---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# isodistrreg: Isotonic Distributional Regression

<!-- badges: start -->
[![Build Status](https://travis-ci.com/AlexanderHenzi/isodistrreg.svg?token=1kfJSpfJj96s5n1DwsiP&branch=master)](https://travis-ci.com/AlexanderHenzi/isodistrreg)
<!-- badges: end -->

Isotonic distributional regression (IDR) is a powerful nonparametric technique
for the estimation of distributions of a binary or numeric response variable
conditional on numeric or ordinal covariates. IDR assumes that there is a
monotone relationship between the response variable and the covariates, where
the partial order on the covariate space can be specified by the user, and has
no tuning parameters. It can be used to generate calibrated probabilistic
weather forecasts from ensemble forecasts and observations, and serve as a
benchmark in many other prediction problems.

## Installation

Download the development version from [GitHub](https://github.com/) with

``` r
# install.packages("devtools")
devtools::install_github("AlexanderHenzi/isodistrreg", build_vignettes = TRUE)
```

or install the package from CRAN
(https://CRAN.R-project.org/package=isodistrreg):

``` r
install.packages("isodistrreg")
```

If the installation fails on Linux because of the error *Vignette re-building
failed*, try installing with `build_vignettes = FALSE` or make sure that 
texinfo is installed (`sudo apt-get install texinfo`).

## Usage Examples

The following basic example illustrates how to use IDR to calibrate forecasts of 
accumulated precipitation.

```{r example, eval=FALSE}
library(isodistrreg)

# Prepare dataset: Half of the data as training dataset, other half for validation.
# Consult the R documentation (?rain) for details about the dataset.
data(rain)
trainingData <- subset(rain, date <= "2012-01-09")
validationData <- subset(rain, date > "2012-01-09")

# Variable selection: use HRES and the perturbed forecasts P1, ..., P50
varNames <- c("HRES", paste0("P", 1:50))

# Partial orders on variable groups: Usual order of numbers on HRES (group '1') and
# increasing convex order on the remaining variables (group '2').
groups <- setNames(c(1, rep(2, 50)), varNames)
orders <- c("comp" = 1, "icx" = 2)

# Fit IDR to training dataset.
fit <- idr(
  y = trainingData[["obs"]],
  X = trainingData[, varNames],
  groups = groups,
  orders = orders
)

# Make prediction for the first day in the validation data:
firstPrediction <- predict(fit, data = validationData[1, varNames])
plot(firstPrediction)

# Use cdf() and qpred() to make probability and quantile forecasts:

## What is the probability of precipitation?
1 - cdf(firstPrediction, thresholds = 0)

## What are the predicted 10%, 50% and 90% quantiles for precipitation?
qpred(firstPrediction, quantiles = c(0.1, 0.5, 0.9))

# Make predictions for the complete verification dataset and compare IDR calibrated
# forecasts to the raw ensemble (ENS):
predictions <- predict(fit, data = validationData[, varNames])
y <- validationData[["obs"]]

## Continuous ranked probability score (CRPS):
CRPS <- cbind(
  "ens" = crps(validationData[, varNames], y),
  "IDR" = crps(predictions, y)
)
apply(CRPS, 2, mean)

## Brier score for probability of precipitation:
BS <- cbind(
  "ens" = bscore(validationData[, varNames], thresholds = 0, y),
  "IDR" = bscore(predictions, thresholds = 0, y)
)
apply(BS, 2, mean)

## Quantile score of forecast for 90% quantile:
QS90 <- cbind(
  "ens" = qscore(validationData[, varNames], quantiles = 0.9, y),
  "IDR" = qscore(predictions, quantiles = 0.9, y)
)
apply(QS90, 2, mean)

## Check calibration using (randomized) PIT histograms:
pitEns <- pit(validationData[, varNames], y)
pitIdr <- pit(predictions, y)

par(mfrow = c(1, 2))
hist(pitEns, main = "PIT of raw ensemble forecasts", freq = FALSE)
hist(pitIdr, main = "PIT of IDR calibrated forecasts", freq = FALSE)
```



## References

Alexander Henzi, Johanna F. Ziegel, and Tilmann Gneiting. Isotonic
Distributional Regression. arXiv e-prints, art. arXiv:1909.03725, Sep 2019. URL 
https://arxiv.org/abs/1909.03725.

The dataframe `precipData_caseStudy.rda` contains all data for the case 
study in the paper.